Two subtle problems with over-representation analysis

### 2024-06-07 analysis/analyse_rnaseq1.Rmd
fig1

# number of sets
lapply(gs,length)

# set size summary
lapply(gs,function(x) {
  y <- unlist(lapply(x, length ))
  summary(y)
} )

# gene coverage
lapply(gs,function(x) {
  y <- unique(unlist(x))
  length(y)
} )

# total number of annotations
lapply(gs, function(x) {
  sum(unlist(lapply(x,length)))
} )

@@

par(mar=c(c(5.1, 7.1, 4.1, 2.1) ))
barplot(t(df),beside=TRUE,horiz=TRUE,las=1,xlim=c(0,1250),
  xlab="no. significant sets", main="Effect of correcting the background bug",
  col=c("gray30","gray80"))
text(df$ns1+50,((1:nrow(df))*3)-1.5,labels=signif(df[,1],3))
text(df$ns2+50,((1:nrow(df))*3)-0.5,labels=signif(df[,2],3))
#abline(v=seq(100,1000,100),lty=2,lwd=0.5,col="gray")
legend("bottomright", inset=.02, legend=c("corrected","original"),
  fill=c("gray80","gray30"), horiz=FALSE, cex=1.1)

@@

# JACCARD
jac <- matrix( res1$jaccard , ncol=9 )
rownames(jac) <- names(d)
colnames(jac) <- names(gs)

@@

barplot(sort(jac2),horiz=TRUE,las=1,xlim=c(0,1),
  main="Jaccard similarity Corrected:Original",xlab="Jaccard index")

>>>>

# number of sets
num <- unlist(lapply(gs,length))
num

# set size summary
sz <- unlist(lapply(gs,function(x) {
  y <- unlist(lapply(x, length ))
  summary(y)
} ))
sz <- t(matrix(sz,ncol=9))
rownames(sz) <- names(gs)
colnames(sz) <- names(summary(c(1,2,3,4)))
sz

# gene coverage
cov <- unlist(lapply(gs,function(x) {
  y <- unique(unlist(x))
  length(y)
} ))
cov

# total number of annotations
tot <- unlist(lapply(gs, function(x) {
  sum(unlist(lapply(x,length)))
} ))
tot

## First test

```{r,initialtest1}

x1 <- readRDS("../dataprep/bulkrna1.Rds")
head(x1)

# remove genes without symbols
x1f <- x1[lapply(strsplit(rownames(x1) , " "),length)>1,]

dim(x1)
dim(x1f)

# about 1000 genes are lost after removing redundant gene names
bg <- unique(sapply(strsplit(rownames(x1f) , " "),"[[",2))
length(bg)

up <-rownames(subset(x1f, log2FoldChange > 0 & padj < 0.05 ))
up <- unique(sapply(strsplit(up," "),"[[",2))
length(up)

dn <-rownames(subset(x1f, log2FoldChange < 0 & padj < 0.05 ))
dn <- unique(sapply(strsplit(dn," "),"[[",2))
length(dn)

# kegg gene set
kegg <- gs[[1]]
kegg <- stack(kegg)
colnames(kegg) <- c("gene","term")
kegg <- kegg[,c(2,1)]
kegg$term <- gsub("KEGG_MEDICUS_","",kegg$term)
kegg$term <- gsub("_"," ",kegg$term)

ora_up <- as.data.frame(enricher(gene = up ,
    universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = kegg,
    pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

  ora_up$es <- gr/br

ora_up$geneID = ora_up$ID = ora_up$Description = NULL

ora_up_kegg <- ora_up

head(ora_up,20) %>%
  kbl(row.names = TRUE, caption="Upregulated KEGGs in SRP128998") %>%
  kable_paper("hover", full_width = F)

# GO gene set
go <- gs[[6]]
go <- stack(go)
colnames(go) <- c("gene","term")
go <- go[,c(2,1)]
go$term <- gsub("^GO","",go$term)
go$term <- gsub("_"," ",go$term)

ora_up <- as.data.frame(enricher(gene = up ,
    universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = go,
    pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

  ora_up$es <- gr/br

ora_up$geneID = ora_up$ID = ora_up$Description = NULL

ora_up_go <- ora_up

head(ora_up,20) %>%
  kbl(row.names = TRUE, caption="Upregulated GOs in SRP128998") %>%
  kable_paper("hover", full_width = F)

Transfer a gene set from GO to KEGG analysis.

```{r,intitialtest2}

pw <- gs[[6]][grep("GOBP_COMPLEMENT_ACTIVATION_CLASSICAL_PATHWAY", names(gs[[6]]))]

pw

pw <- stack(pw)[,c(2,1)]
colnames(pw) <- c("term","gene")
kegg2 <- rbind(kegg,pw)

ora_up <- as.data.frame(enricher(gene = up ,
    universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = kegg2,
    pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

  ora_up$es <- gr/br

ora_up$geneID = ora_up$ID = ora_up$Description = NULL

ora_up_kegg2 <- ora_up

head(ora_up,20) %>%
  kbl(row.names = TRUE, caption="Upregulated KEGGs in SRP128998 with Complement") %>%
  kable_paper("hover", full_width = F)

message("GO")
ora_up_go[rownames(ora_up_go)=="BP COMPLEMENT ACTIVATION CLASSICAL PATHWAY",]

message("KEGG + complement")
ora_up_kegg2[rownames(ora_up_kegg2)=="GOBP_COMPLEMENT_ACTIVATION_CLASSICAL_PATHWAY",]

Fix KEGG analysis.

```{r,keggfix1}

bgdf <- data.frame("background",bg)
colnames(bgdf) <- c("term","gene")
keggfix <- rbind(kegg,bgdf)

ora_up <- as.data.frame(enricher(gene = up ,
    universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = keggfix,
    pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

  ora_up$es <- gr/br

ora_up$geneID = ora_up$ID = ora_up$Description = NULL

ora_up_keggfix <- ora_up

head(ora_up,20) %>%
  kbl(row.names = TRUE, caption="Upregulated KEGG fixed in SRP128998 with Complement") %>%
  kable_paper("hover", full_width = F)

message("Original")
ora_up_kegg[which(rownames(ora_up_kegg)=="REFERENCE CLASSICAL PATHWAY OF COMPLEMENT CASCADE C4 C2 TO C3 CONVERTASE FORMATION"),]

message("Fixed")
ora_up_keggfix[which(rownames(ora_up_keggfix)=="REFERENCE CLASSICAL PATHWAY OF COMPLEMENT CASCADE C4 C2 TO C3 CONVERTASE FORMATION"),]

@@

par(mar=c(c(5.1, 7.1, 4.1, 2.1) ))
barplot(t(df),beside=TRUE,horiz=TRUE,las=1,xlim=c(0,1250),
  xlab="no. significant sets",
  col=c("gray30","gray80"))
text(df$ns1+50,((1:nrow(df))*3)-1.5,labels=signif(df[,1],3))
text(df$ns2+50,((1:nrow(df))*3)-0.5,labels=signif(df[,2],3))
#abline(v=seq(100,1000,100),lty=2,lwd=0.5,col="gray")
legend("bottomright", inset=.02, legend=c("corrected","original"),
  fill=c("gray80","gray30"), horiz=FALSE, cex=1.1)

@@

png("fig1_ORAsig.png")
par(mar=c(c(5.1, 7.1, 2.1, 2.1) ))
barplot(t(df),beside=TRUE,horiz=TRUE,las=1,xlim=c(0,1250),
  xlab="no. significant sets",
  col=c("gray30","gray80"))
text(df$ns1+50,((1:nrow(df))*3)-1.5,labels=signif(df[,1],3))
text(df$ns2+50,((1:nrow(df))*3)-0.5,labels=signif(df[,2],3))
#abline(v=seq(100,1000,100),lty=2,lwd=0.5,col="gray")
legend("bottomright", inset=.02, legend=c("corrected","original"),
  fill=c("gray80","gray30"), horiz=FALSE, cex=1.1)

mylabels <- gsub("$","%",gsub("^","+",signif(((df$ns2/df$ns1)-1)*100,3)))
text(df$ns2+160,((1:nrow(df))*3)-1,labels=mylabels,cex=1.1)
dev.off()

# JACCARD
jac <- matrix( res1$Jaccard , ncol=9 )
rownames(jac) <- names(d)
colnames(jac) <- names(gs)

@@

par(mar=c(c(5.1, 7.1, 2.1, 2.1) ))
barplot(sort(jac2),horiz=TRUE,las=1,xlim=c(0,1), xlab="Jaccard index")

par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
plot(cov,jac2,xlim=c(0,30000),ylim=c(0,1),pch=19,col="lightgray",cex=3,
  xlab="gene coverage",ylab="Jaccard index")
text(cov,jac2,labels=names(cov))
mylm <- lm(jac2~cov)
abline(mylm,lty=2,lwd=2)
SLOPE=unname(signif(mylm$coefficients[2],3))
INT=unname(signif(mylm$coefficients[1],3))
PVAL=signif(summary(mylm)$coefficients[4],3)
RSQ=signif(summary(mylm)$r.squared,3)
HEADER=paste("Rsq =",RSQ,"; p-val =",PVAL)
mtext(HEADER)

png("fig1_ORAjac.png")
par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
plot(cov,jac2,xlim=c(0,30000),ylim=c(0,1),pch=19,col="lightgray",cex=3,
  xlab="gene coverage",ylab="Jaccard index")
text(cov,jac2,labels=names(cov))
mylm <- lm(jac2~cov)
abline(mylm,lty=2,lwd=2)
SLOPE=unname(signif(mylm$coefficients[2],3))
INT=unname(signif(mylm$coefficients[1],3))
PVAL=signif(summary(mylm)$coefficients[4],3)
RSQ=signif(summary(mylm)$r.squared,3)
HEADER=paste("Rsq =",RSQ,"; p-val =",PVAL)
mtext(HEADER)
dev.off()

### 2024-06-07 manuscript/manuscript.Rmd
fig1

[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352),
[Mark Ziemann<sup>1,2</sup>*](https://orcid.org/0000-0002-7688-6974)

@@

Functional enrichment analysis involves the summarisation of omics data to reflect
biological differences, and has become one of the most popular methods in bioinformatics.
Enrichment analysis software packages have been collectively cited 131,332 times as of
late 2019 [@Xie2021-ch].
The algorithms used by these tools can largely be divided into two groups, functional
class scoring (FCS) and over-representation analysis (ORA).
In FCS, all detected genes are ranked by their degree of differential expression followed
by a statistical test for enrichment of gene categories at either extreme of the
ranked list.
ORA involdes the selection of genes of interest, followed by a test to ascertain whether
certain functional categories are over-represented in the selected genes.

According to a study that used simulated differential expression data, FCS accuracy was
superior to ORA over a range of experimental designs [@Kaspi2020-rn].
Despite this performance difference, a survey of articles published in 2019 showed that
seven of the top eight most popular tools were based on ORA [@Wijesooriya2022-fw].
The discrepancy between performance and popularity likely arose due to the relative ease of
conducting ORA as compared to FCS.
A minimalist ORA might involve pasting a single list of genes into a text box on a website,
where the results appear nearly instantaneously.
On the other hand, FCS typically involves downloading specific softwares and dealing with a
dataset that represents every detected gene, typically ~20,000 rows and ensuring that the
format from upstream tools can be shoehorned to meet the requirements of FCS packages.

Since it's initial development in 1999 [@Tavazoie1999-bu], ORA has proliferated widely,
becoming part of many websites and software packages.
The most popular ORA website is DAVID [@Dennis2003-ar;@Huang2009-gi;@Sherman2022-jn],
followed by Ingenuity Pathway Analysis, a commercial software package provided by QIAGEN Inc.
ORA has been implemented into popular R/Bioconductor packages including `ClusterProfiler`
[@Yu2012-jj],`limma` [@Ritchie2015-oz] and `GOseq` [@Young2010-iw].

@@

The focus of this work is to raise awareness to two existing problems in ORA as implemented
by some popular enrichment tools.
The first problem is that genes belonging to the background list are removed from the analysis
entirely if they are not annotated as belonging to any functional categories.
This results in the background list appearing smaller than it should, leading to an
underestimation of fold enrichment scores and significance values.
In principle, this problem would differentially affect analyses involving different gene set
libraries, with smaller libraries such as Kyoto Encyclopedia of Genes and Genomes (KEGG)
[@Kanehisa2023-ni], which describes 2,727 genes affected more severely as compared to larger
libraries like Gene Ontology [@The_Gene_Ontology_Consortium2023-xp] which describes 19,428
genes (figures obtained from MSigDB [@Liberzon2015-cs]).

The second problem is that adjusting p-values for multiple testing, also known as
false-discovery correction is implemented improperly.
ORA tools should implement a threshold by which a pathway could be considered as detected.
Ideally, a pathway should be considered detected based on the presence of a predetermined
number of member genes in the background set.
However some tools use the foreground set to define a pathway as detected.
This is problematic, as the act of trying to find an intersect between a foreground gene list
and a pathway is in itself a test, even if there are no common genes.
This results in pathways being discarded from the analysis if no common genes are found in
the foreground, despite a large number of those genes being present in the background list.
This leads to FDR values appearing lower than they should.
We expect this problem to have a more severe effect when the foreground list is fairly small.

Here, we show the effect of these two problems on real RNA-seq based enrichment analysis
results and demonstrate these two issues lead to poor performance using simulated expression
data.
Finally we classify popular tools regarding these two problems and provide general
recommendations for end users.

## A systematic assessment of reproducibility in enrichment analysis

@@

@@

@@

* The full protocol can be found at protocols.io
(https://www.protocols.io/view/a-recipe-for-extremely-reproducible-enrichment-ana-j8nlkwpdxl5r/v2) @Ziemann2023-kz.

* Instructional video guides are available on YouTube
(https://www.youtube.com/playlist?list=PLAAydBPtqFMXDpLa796q7f7W1HK4t_6Db).

>>>>

[Mark Ziemann<sup>1,2</sup>*](https://orcid.org/0000-0002-7688-6974),
[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352)

@@

Over-representation analysis (ORA) is a type of functional enrichment analysis (FEA) that
involves the summarisation of omics data to reflect biological differences.
Generally speaking, FEA has become one of the most popular methods in bioinformatics,
collectively accumulating 131,332 citations as of late 2019 [@Xie2021-nv].

ORA involdes the selection of genes of interest, followed by a test to ascertain whether
certain functional categories are over-represented in the selected genes.
Since it's initial development in 1999 [@Tavazoie1999-bu], ORA has proliferated widely,
becoming part of many websites and software packages.
The most popular ORA website is DAVID [@Dennis2003-ar;@Huang2009-gi;@Sherman2022-jn],
followed by Ingenuity Pathway Analysis, a commercial software package provided by QIAGEN Inc.
ORA has been implemented into popular R/Bioconductor packages including `ClusterProfiler`
[@Yu2012-jj;@Wu2021-wy], `limma` [@Ritchie2015-oz] and `GOseq` [@Young2010-iw].

Another popular approach to FEA is functional class scoring (FCS).
In FCS, all detected genes are ranked by their degree of differential expression followed
by a statistical test for enrichment of gene categories at either extreme of the
ranked list [@Subramanian2005-no].

According to a study that used simulated differential expression data, FCS accuracy was
superior to ORA over a range of experimental designs [@Kaspi2020-rn].
Despite this performance difference, a survey of articles published in 2019 showed that
seven of the top eight most popular tools were based on ORA [@Wijesooriya2022-fw].
The discrepancy between performance and popularity is likely due to the relative ease of
conducting ORA as compared to FCS.
A minimal ORA might involve pasting a single list of genes into a text box on a website,
where the results appear nearly instantaneously.
On the other hand, FCS typically involves installing specific softwares and dealing with a
dataset representing every detected gene (~20,000 rows) and ensuring that the file
format from upstream tools can be compatible with FCS packages.

@@

The focus of this work is to raise awareness to two existing problems in some implementations
of ORA used by popular enrichment tools.
The first problem is that genes belonging to the background list are removed from the analysis
entirely if they are not annotated as belonging to any functional categories.
This results in the background list appearing smaller than it should, leading to an
underestimation of fold enrichment scores and significance values.
In principle, this problem would differentially affect analyses involving different gene set
libraries, with smaller libraries such as Kyoto Encyclopedia of Genes and Genomes (KEGG)
[@Kanehisa2023-ni], which describes 2,727 genes affected more severely as compared to larger
libraries like Gene Ontology [@The_Gene_Ontology_Consortium2023-xp] which describes 19,428
genes (figures obtained from MSigDB [@Liberzon2015-cs]).

The second problem is that adjusting p-values for multiple testing, also known as
false-discovery correction is implemented improperly.
ORA tools should implement a threshold by which a pathway could be considered as detected.
Ideally, a pathway should be considered detected based on the presence of a predetermined
number of member genes in the background set.
However some tools use the foreground set to define a pathway as detected.
This is problematic, as the act of trying to find an intersect between a foreground gene list
and a pathway is in itself a test, even if there are no common genes.
This results in pathways being discarded from the analysis if no common genes are found in
the foreground, despite a large number of those genes being present in the background list.
This artificially reduces the number of tests conducted and makes FDR values appear lower
than they should, potentially raising the rate of false positives.
We expect this problem to have a more severe effect when the foreground list is fairly small.

Here, we define the effect of these two problems on real RNA-seq based enrichment analysis
results and demonstrate how these two issues impact performance using simulated expression
data.
Finally we classify popular tools regarding these two problems and provide general
recommendations for end users.

@@

### Expression data preparation

To test the effect of these two problems on RNA-seq based enrichment analysis, data from seven
transcriptomic studies were downloaded from DEE2 using the getDEE2 R/Bioconductor package
[@Ziemann2019-ux].
Raw data for these seven studies are available from NCBI Sequence Read Archive under the
accession numbers SRP128998, SRP038101, SRP037718, SRP096177, SRP097759, SRP253951 and
SRP068733 [@Felisbino2021-hc; @Lund2014-ds; @Rafehi2014-rv; @Keating2014-ab;
@Blanco-Melo2020-hh; @Rafehi2017-qn].
For each dataset, kallisto counts were aggregated to the gene level.
Genes with fewer than 10 reads per sample on average were removed from downstream analysis.
Remaining genes passing this selection were included in the background gene set.
Differential expression analysis was conducted with DESeq2 v1.42.0 [@Love2014-wg], with
genes identified by their Ensembl identifiers.
Gene symbols were then fetched using biomaRt v2.58.0, based on Ensembl version 112 [@Durinck2009-dr].

### Examining the background problem using real transcriptome data

Human gene sets were obtained from MSigDB v2023.2 [@Liberzon2015-cs], and included
KEGG Medicus, Reactome, Wikipathways, MicroRNA targets from miRdb, transcription factor
targets from GTRD, Gene Ontology (GO), Human Phenotype Ontology (HPO), Cell markers and
Hallmark pathways [REFS].
For an initial demonstration of the background problem, differentially expressed genes with
FDR<0.05 were subset, with separate lists for up and downregulated genes.
These gene lists were subjected to ORA using the `enricher` function of clusterProfiler
v4.10.0, using a minimum set size of 5 and a maximum set size of 500,000.

For one test, a GO gene set called "Complement activation classical pathway" was added
to the KEGG database to test whether fold enrichment and p-values were affected.

To systematically assess the impact of the background problem, we conducted ORA with
`enricher` using the different gene set libraries described above for the seven RNA-seq
datasets and nine gene set libraries.
`enricher` was used with the same parameters as above, and we devised a simple workaround
to the background problem, which is to append the entire background list as a gene set to the
library.
This forces `enricher` to retain all background genes.
The ORA analyses were then filtered for significant sets using a FDR threshold of 0.05.
Jaccard index was used to compare original and corrected analyses.

### Examining the FDR problem

@@

### Understanding the background problem

In the analysis of dataset 1, the background list contained 14,144 genes and the foreground
list consisted of 1,469 genes.
After ORA with KEGG sets, the background size reported by clusterProfiler was only 2,120.
With GO sets, the background was reported to be 12,365.
These numbers correspond to the number of genes in the background list that are annotated with
one or more functions in the respective gene set libraries.

We used a workaround to eliminate the background problem and compared the results to standard
analysis.
This was repeated for seven datasets and the nine gene set libraries.
The mean number of significant results was calculated for each of the gene set libraries
before and after correcting the problem (**Figure 1A**).
Each of the gene set libraries showed an uplift when correcting the background problem.
The smallest increase was observed in Cellmarkers gene library (11.7%), while the largest
increase was observed for HPO (229%).
The similarity between results, as quantified with the Jaccard similarity was highest for
Cellmarkers and GTRD at 0.88 and 0.76 respectively, and the lowest similarity was recorded for
HPO and KEGG with 0.20 and 0.48 respectively.
Observed Jaccard scores correlated with the number of genes annotated to one or more functional
categories in the gene set library (**Figure 1B**).

![**Figure 1A. Effect of correcting the background problem.** Figures represent the mean number
of statistically significant sets over seven independent datasets detected with clusterProfiler
with and without correcting the background problem.
The percent increase in significant sets are shown.](../analysis/fig1_ORAsig.png "Figure 1A
shows that correcting the background bug improves the number of significant sets between 12 and
229%.")

![**Figure 1B. Impact of background problem is worse for smaller gene set libraries.** Mean
Jaccard similarity index was calculated for original and corrected ORA for seven independent
datasets. The number of genes represented in one or more functional sets is shown on the x-axis,
and the mean Jaccard index is shown on the y-axis.](../analysis/fig1_ORAjac.png "Figure 1B
shows that the background bug impacts smaller gene sets libraries like KEGG and HPO more
severely.")

### Understanding the FDR problem

@@
